INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-22 chapters/background.tex
writing section 3

\subsection{Related Projects}

@@

Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

>>>>

\subsection{Related Projects}
\label{sect:RELATED}

@@

The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

### 2016-10-22 chapters/evaluation.tex
fix to make the paper compile
finished introduction
writing section 3

\section{Evaluation}

@@

We experimented with import of XML, JavaScript, Python, Ruby, C#, Swift, MySQL and some other languages.

>>>>

\section{Evaluation}
\label{sect:EVAL}

\PP{Zhodnotit jak se podarilo dosahnout automatizace. How much we fulfilled our goal of maximal possible automation.}

@@

We experimented with import of XML, JavaScript, Python, Ruby, C\#, Swift, MySQL and some other languages.

### 2016-10-22 chapters/language_definition.tex
related projects
writing section 3

\section{Creating Language Definition in MPS}
\label{sect:LANGDEF}

From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
\begin{itemize}
	\item It is a very widely used grammar notation.
	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation). This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
\end{itemize}

We use ANTLRv4 grammars as inputs (and work with them).
	https://github.com/antlr/antlr4/blob/master/doc/index.md
	https://github.com/antlr/grammars-v4

Unlike some of the related projects, we did not add any new features into the grammar notation.
We tried to import grammars as they are.

The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.

To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.

Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
In particular, we did not port the ANTLR notation as an MPS language.
In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).

During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.

Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
\PV{Dalsi veta asi zbytecna}
We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.

Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.

\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
There are more ways how to approach the problem of a grammar import.

Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion. In particular, MPS UI allows the user to perform only changes that result in valid output.
Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.

Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
We blame all this on the project being in its early stage of development.

In the rest of this section, first we define the language that we will use throughout the paper as the running example.
Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
At some places, we discuss important possible alternatives (as there are several possible approaches).

@@

We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).

@@

@@

The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
The structure of the MPS language is somewhat similar to the grammar.
This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.

@@

@@

The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
In particular, the grammar says nothing about indentation, line breaks, and another formatting.

>>>>

\section{Constructing Languages in MPS}
\label{sect:LANGDEF}

The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

The process of MPS language construction, as performed by INGRID, consists of four steps --- parsing of the input grammar, followed by definition of the essential aspects (Structure, Editor, and TextGen).
Each of the steps 2-4 is exclusively and fully responsible for one aspect.

INGRID is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by INGRID.

Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.

% TODO continue from here

@@

We illustrate the main steps of the INGRID method on the example of a simplified XML language, which is defined below (in Figure XY).

@@

\todo{We import/generate the complete/full language structure fully automatically.}

@@

The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
The structure of the MPS language is somewhat similar to the grammar.
This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.

As we already indicated in the introduction, the main problem/difficulty/challenge related to Structure (creating the structure) is that a grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
Challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component).
There is a challenge/issue that is related to construction of the structure aspect.
Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
In the rest of this section, we describe our solution to this problem/challenge.

@@

\todo{Definition of projectional editors is only partially automated. We do a lot of mundane and time-consuming work connected to creating projectional editors.}

@@

The main problems/challenges in the process of definition of projectional editors, also/again related to input grammars, are code layout is whitespace (editor aspect).
As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language (the description of a language in the form of a grammar) does not hold any information about the code layout.
The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
In particular, the grammar says nothing about indentation, line breaks, and another formatting.
